# Phase II trials

```{r}
#| echo: false
#| warning: false
#| results: hide

library(splitstackshape)
library(tidyverse)
library(extraDistr)
library(data.table)
```

Phase II trials are often single-arm trials in early phases but can also be multi-arm (non-)randomized controlled trials in later phases @Mossop2022. The aim of early phase-II designs is to assess efficacy of a new treatment measured by clinical response in a small population of 30-100 patients. Early phase II trials can be designed as two- or three stage trials using frequentist approaches like Simon's two stage design. In this chapter we use a Bayesian predictive probability approach to analyse an early phase-II trial @Lee2008 @Sambucini2021. The aim is to compare two different decision rules for early stopping the trial: One uses the posterior distribution given the interim data, the second one uses the predictive distribution given the interim data and future events. 


## Single arm design

Let $p$ the true toxicity rate and $\pi \sim Beta(a,b)$ be a design prior on $p$. Our null hypothesis of interest is $H_0: p \leq p_0$ versus $H_1: p>p_0$, that is, we compare, say, the clinical response $p$ toa prespecified threshold $p_0$. Suppose that we plan interim looks $j\geq 1$ after $N_j$ patients were recruited and $r_j$ toxicity events occurred. Under the prior $\pi$ the posterior distribution is beta distributed $p|r_j \sim Beta(a+r_j, b+N_j-r_j)$. Suppose that we plan to recruit $N_{max}$ patients. Let $Y_j$ be the number of toxicities in future $m_j=N_{max}-N_j$ patients at interim look $j\geq 1$. Under the prior $\pi$, $Y_j$ is beta-binomial distributed $Y_j \sim Betabinomal(i|m_j, a+r_j, b+N_j-r_j)$ and $p|r_j,y_j \sim Beta(a+r_j+y_j, b+N_{max}-N_j-r_j-y_i)$.

------------------------------------------------------------------------

**Working example**

Suppose that the goal of a phase-II trial is to evaluate safety of new chemotherapy treatment investigating the discontinuation rate $p$ as primary endpoint. Discontinuation is defined based on toxicity grading from the US National Cancer Institute Common Terminology Criteria for Adverse Events (CTCAE).

------------------------------------------------------------------------

### Decision rules for stopping the trial

**Early stopping using the posterior distribution**

The decision to the stop the trial early is based on the decision rule $P(p> p_0|N_j)>\theta_T$, that is, the posterior distribution given the interim data.  If the trial is not stopped early we recruit more patients until $N_{max}$ is reached.

**Early stopping using the predictive distribution**

Let $T_{j,i}:=P(p> p_0|N_j, Y_j=i)$, that is, the posterior distribution given the interim data and future events. Note that because $Y_j$ are not observed events $T_{j,i}$ are random variables. The predictive probability for a successful (safe) trial is

$$
PP_j=\sum_{i\leq m} P(Y_j=i|r_j)\cdot I(T_{j,i}>\theta_T), \quad j\geq 1.
$$
This is the expectation of $T_{j,i}$ given the interim data $j\geq 1$ @Lee2008. If $PP_j>\theta_S$ then we stop the trial because of safety concerns, otherwise we recruit more patients until $N_{max}$ is reached. Usually, $\theta_S$ is chosen to be high (>0.8) @Sambucini2021.

------------------------------------------------------------------------

**Working example**

Based on prior evidence $p_0$ is set to 0.2. This threshold is based on a result from a previous study where the treatment discontinuation was estimated with 0.17 with 95% confidence intervals (0.10-0.27). Below we show the prior distributions for different parameters $a=1$, $b=4$ (weight of 5 patients) and $a=2$, $b=8$ (weight of 10 patients). The mean of the prior is in both cases 0.2.


```{r}
#| echo: true
#| warning: false

# Prior information
a_0 <- 2
b_0 <- 8

x <- seq(0.001,0.999,0.001)

data_prior <- data.frame(x, y=dbeta(x, a_0, b_0), type=1)

# Prior information
a_0 <- 1
b_0 <- 4

data_prior <- bind_rows(data_prior, data.frame(x, y=dbeta(x, a_0, b_0), type=2))

data_prior$type <- factor(data_prior$type, levels=1:2, 
                          labels=c("a=0.4, b=0.6", "a=1, b=4"))
ggplot(data_prior %>% filter(y<5), aes(x=x, y=y, linetype=type, group=type))+
  geom_line()+theme_bw()+scale_x_continuous(breaks=seq(0,1,0.2))+
  theme(panel.grid = element_blank(), legend.position = "bottom")+
  scale_linetype("Prior parameters")+geom_vline(xintercept=0.2, colour="red", linetype=1)+
  geom_vline(xintercept=0.2, colour="red", linetype=2)+
  ylab("Density")+xlab("")+ggtitle("Different beta prior parameter choices")+
  labs(caption=c("Red vertical lines indicate mean values of priors."))
```


```{r}
#| echo: true
#| warning: false

p_0_max <- 0.2

# Prior information
a_0 <- 0.4
b_0 <- 0.6

n <- 35
r <- 0:n

n_max <- 35
n <- seq(10, n_max, 5)
r <- 0:n_max
m <- n_max-r
x <- seq(0,1,0.001)

theta_T <- 0.75

data <- expand.grid(n=n, r=r) %>% arrange(n) %>% filter(n>=r) %>% mutate(i=n_max-n+1)
data <- expandRows(data, count=3)
data <- data %>% group_by(n, r) %>% mutate(m=n_max-n, ind=1, i=cumsum(ind)-1, ind=NULL)

data <- data %>% mutate(cond_postprob=dbbinom(i, size=m, alpha=a_0+r, beta=b_0+n-r), 
                        Ti=1-pbeta(p_0_max, a_0+r+i, b_0+n_max-r-i), 
                        ind=ifelse(Ti>theta_T, 1, 0), n_max)

data <- data %>% select(n_max, n, r, m, i, cond_postprob, Ti, ind)
```

Suppose we plan a phase-II with a maximum of $N_{max}=35$ patients. Let us consider the situation where we have reached an initial $N_1=10$ with $r_1=2$ observed toxicities and we have to decide whether to stop or continue the trial. Specifically, we calculate for each potential future toxicity from $m_1$ patients and decide whether $T_{i,1}>\theta_T=0.75$. The column *cond_postprob* in the following table is $p|r_1,y_i$ (because it is conditional on the future outcome $Y_{j,i}=i$) and the column $Ti$ corresponds to $T_{1,i}$.

```{r}
#| echo: true
#| warning: false

data %>% filter(n==10, r==2, i<=15)
```

Based on this information we calculate $PP_1$ and decide based on the threshold $\theta_S=0.9$ whether to stop the trial early or to accrual more patients. For the decision rule based on the posterior probability we calculate the decision to stop the trial using the posterior distribution and use the threshold $\theta_T=0.75$.

```{r}
#| echo: true
#| warning: false

threshold_safety <- 0.9

## Early stopping using predictive distribution
data_pp <- data %>% group_by(n, r) %>% summarise(pp=round(sum(cond_postprob*ind),6), 
                                                 stop_pp=ifelse(pp>threshold_safety, 1, 0))

## Early stopping using posterior distribution
data_pp <- data_pp %>% 
  mutate(stop_postdist=ifelse(1-pbeta(p_0_max, a_0+r, b_0+n-r)>theta_T, 1, 0))

data_pp %>% filter(n==10)
```

For each interim look $j\geq1$ we can then calculate stopping of safety boundaries of the trial. Here we assumed that we recruit 5 additional patients if we do not stop the trial early.

```{r}
#| echo: true
#| warning: false

data_pp_long <- pivot_longer(data_pp, cols=4:5)

stop_boundaries <- data_pp_long %>% filter(value==1) %>%
  group_by(name, n) %>% summarise(r=min(r), stop=NULL)

stop_boundaries$name <- factor(stop_boundaries$name, 
                               levels=c("stop_postdist", "stop_pp"),
                               labels=c("Posterior distribution",
                                        "Predictive distribution"))

ggplot(stop_boundaries, aes(x=n, y=r, linetype=name))+geom_step()+
  scale_y_continuous(limits=c(0, max(stop_boundaries$r)), 
                     breaks=0: max(stop_boundaries$r))+
  ylab("Number of toxic events")+xlab("Number of patients at interim analysis")+
  theme_bw()+theme(panel.grid = element_blank())+
  ggtitle("Stopping for safety boundaries")+
  scale_linetype("")
```

------------------------------------------------------------------------

### Simulation study

#### Prior with a=0.4, b=0.6

We set up a simulation study to quantify operations characteristics (probability of early stopping, the probability to declare toxicity and expected number of patients and toxicities) with the following parameters

- Beta prior parameters: a=0.4, b=0.6
- True $p$: 0.2, 0.25, 0.3
- $p_0$: 0.2
- $N_0$: 10
- Cohort increase: 5
- $N_{max}$: 30, 35, 40
- $\theta_T$: 0.75, 0.9
- $\theta_S$: 0.8, 0.9, 1
- Number of simulation runs: 1000

Stopping decision rules:

- Posterior distribution: Stop early if $P(p> p_0|N_j)>\theta_T$,
- Safety threshold: Stop early if $T_{j,i}>\theta_T$ and $PP_j>\theta_S$.

Note that the type-I and type-II errors correspond to the proportion of trials which stopped early. If the maximal number of patients was reached the trial cannot be stopped early. 

```{r}
#| echo: false
#| warning: false
#| cache: true

set.seed(52359)

n_sim <- 1000

a_0 <- 0.4
b_0 <- 0.6

p_0_max <- 0.2

true_p_0_values <- c(0.2, 0.25, 0.3)
n_max_values <- c(30, 35, 40)
theta_T_values <- c(0.75, 0.9)
safety_threshold_values <- c(0.8, 0.9, 1)
initial_n <- 10
n_increase <- 5

res_sim <- c()
res_overall1 <- c()

for (safety_threshold in safety_threshold_values) {
  for (true_p0 in true_p_0_values) {
    for (theta_T in theta_T_values) {
      for (n_max in n_max_values) {
        for (i in 1:n_sim) {
          
          data_core <- rbinom(initial_n, size=1, prob=true_p0)
          
          n <- initial_n
          
          while (n<(n_max+1)) {
  
            data_sim_res <- data.frame(n=length(data_core), r=sum(data_core), m=n_max-length(data_core)+1)
            data_sim <- data.frame(n=rep(n, n_max-length(data_core)+1), r=rep(sum(data_core), n_max-length(data_core)+1))
            data_sim$i <- (1:nrow(data_sim))-1
            data_sim$m <- n_max-data_sim$n
            data_sim$cond_prob <- dbbinom(data_sim$i, size=data_sim$m, alpha=a_0+data_sim$r, beta=b_0+data_sim$n-data_sim$r)
            data_sim$ind <- ifelse(1-pbeta(p_0_max, a_0+data_sim$r+data_sim$i, b_0+n_max-data_sim$r-data_sim$i)>theta_T, 1, 0)
            data_sim <- as.data.table(data_sim)
            data_sim_res$pp <- as.numeric(data_sim[, .(sum(cond_prob*ind))])
            data_sim_res$true_p0 <- true_p0
            data_sim_res$theta_T <- theta_T
            data_sim_res$n_max <- n_max
            data_sim_res$n_sim <- i
            data_sim_res$safety_threshold <- safety_threshold
  
            if (data_sim_res$pp<=safety_threshold) {
              n <- n+n_increase
              data_core <- c(data_core, rbinom(n_increase, size=1, prob=true_p0))
              
            }
            else {
              n <- n_max
              n <- n+n_increase
            }
            res_sim[[i]] <- data_sim_res
          }
        }
        res_overall1 <- bind_rows(res_overall1, res_sim)
      }
    }
  }
}

res_sim <- c()
res_overall2 <- c()

for (true_p0 in true_p_0_values) {
  for (theta_T in theta_T_values) {
    for (n_max in n_max_values) {
      for (i in 1:n_sim) {
  
        data_core <- rbinom(initial_n, size=1, prob=true_p0)
        
        n <- initial_n
        
        while (n<(n_max+1)) {
          
          data_sim <- data.frame(n=length(data_core), r=sum(data_core))
          data_sim$postdist <- 1-pbeta(p_0_max, a_0+data_sim$r, b_0+data_sim$n-data_sim$r)
          data_sim$stop <- ifelse(data_sim$postdist>theta_T, 1, 0)
          data_sim$true_p0 <- true_p0
          data_sim$theta_T <- theta_T
          data_sim$n_max <- n_max
          data_sim$n_sim <- i
          

          if (data_sim$stop==0) {
            n <- n+n_increase
            data_core <- c(data_core, rbinom(n_increase, size=1, prob=true_p0))
            
          }
          else {
            n <- n_max
            n <- n+n_increase
          }
          res_sim[[i]] <- data_sim
        }
      }

      res_overall2 <- bind_rows(res_overall2, res_sim)
    }
  }
}

res_overall2$pp <- res_overall2$postdist
res_overall2$safety_threshold <- res_overall2$theta_T
res_overall2$ind <- 1

res_overall <- bind_rows(res_overall1, res_overall2)
```

```{r}
#| echo: false
#| warning: false
#| cache: false
#| fig-height: 10
#| fig-width: 8

res_overall <- res_overall %>% mutate(stop_safety=ifelse(pp>safety_threshold, 1, 0), stop_tox=ifelse(pp>safety_threshold, 1, 0))
res_overall$safety_threshold[res_overall$ind%in%1] <- 999
# Trials who reach N_max cannot stop early
res_overall$stop_safety[res_overall$n==res_overall$n_max] <- 0

res_overall$obs_tox <- round(res_overall$r/res_overall$n, 3)

res_agg <- res_overall %>%  group_by(n_max, theta_T, true_p0, safety_threshold) %>% summarise(stop_safety=mean(stop_safety), stop_tox=mean(stop_tox), obs_tox=mean(obs_tox), exp_n=mean(n), exp_r=mean(r))
res_agg$theta_T <- factor(res_agg$theta_T, levels=c(0.75,0.8,0.9), labels=paste0("Theta_T: ", c(0.75,0.8,0.9)))
res_agg$true_p0 <- factor(res_agg$true_p0, levels=c(0.2,0.25,0.3), labels=c(paste0("True p=", c(0.2), " (H0)"), paste0("True p=", c(0.25,0.3), " (H1)")))
res_agg$safety_threshold <- factor(res_agg$safety_threshold, levels=c(0.8,0.9,1, 999), labels=c(paste0("Safety threshold: ", c(0.8,0.9,1)), "Posterior distribution"))

res_agg$n_max <- as.factor(res_agg$n_max)

fig <- ggplot(res_agg, aes(x=n_max, y=stop_safety, shape=safety_threshold)) + geom_point(position=position_dodge2(width = 0.2)) + facet_wrap(~true_p0+theta_T, ncol=2) + theme_bw() + theme(panel.grid.minor = element_blank())+geom_hline(yintercept=0.2, linetype="dashed") + geom_hline(yintercept=0.8, linetype="dashed") + scale_y_continuous(breaks=c(0,0.2,0.5,0.8, 1), limits=c(0,1))+xlab("N_max")+ylab("Probability") + scale_shape("Stopping decision rule")+ggtitle("Probability of early stopping")
fig

fig <- ggplot(res_agg, aes(x=n_max, y=stop_tox, shape=safety_threshold)) + geom_point(position=position_dodge2(width = 0.2)) + facet_wrap(~true_p0+theta_T, ncol=2) + theme_bw() + theme(panel.grid.minor = element_blank())+geom_hline(yintercept=0.2, linetype="dashed") + geom_hline(yintercept=0.8, linetype="dashed") + scale_y_continuous(breaks=c(0,0.2,0.5,0.8, 1), limits=c(0,1))+xlab("N_max")+ylab("Probability") + scale_shape("Stopping decision rule")+ggtitle("Probability of declaring toxicity")
fig
```

```{r}
#| echo: false
#| warning: false
#| cache: false
#| fig-height: 5
#| fig-width: 8

fig <- ggplot(res_agg %>% filter(true_p0=="True p=0.2 (H0)"), aes(x=n_max, y=exp_n, shape=safety_threshold)) + geom_point(position=position_dodge2(width = 0.2)) + facet_wrap(~theta_T) + theme_bw() + theme(panel.grid.minor = element_blank())+xlab("N_max")+ylab("Expected N under H0") + scale_shape("Stopping decision rule")
fig+ggtitle("Expected N under H0")

# library(DT)
# datatable(res_agg, rownames = F, filter = 'top')
```

#### Prior with a=1, b=4

We use the same simulation approach as in the previous subsection but using a a Beta prior with parameters a=1 and b=4.

```{r}
#| echo: false
#| warning: false
#| cache: true

set.seed(52359)

a_0 <- 1
b_0 <- 4

res_sim <- c()
res_overall1 <- c()

for (safety_threshold in safety_threshold_values) {
  for (true_p0 in true_p_0_values) {
    for (theta_T in theta_T_values) {
      for (n_max in n_max_values) {
        for (i in 1:n_sim) {
          
          data_core <- rbinom(initial_n, size=1, prob=true_p0)
          
          n <- initial_n
          
          while (n<(n_max+1)) {
  
            data_sim_res <- data.frame(n=length(data_core), r=sum(data_core), m=n_max-length(data_core)+1)
            data_sim <- data.frame(n=rep(n, n_max-length(data_core)+1), r=rep(sum(data_core), n_max-length(data_core)+1))
            data_sim$i <- (1:nrow(data_sim))-1
            data_sim$m <- n_max-data_sim$n
            data_sim$cond_prob <- dbbinom(data_sim$i, size=data_sim$m, alpha=a_0+data_sim$r, beta=b_0+data_sim$n-data_sim$r)
            data_sim$ind <- ifelse(1-pbeta(p_0_max, a_0+data_sim$r+data_sim$i, b_0+n_max-data_sim$r-data_sim$i)>theta_T, 1, 0)
            data_sim <- as.data.table(data_sim)
            data_sim_res$pp <- as.numeric(data_sim[, .(sum(cond_prob*ind))])
            data_sim_res$true_p0 <- true_p0
            data_sim_res$theta_T <- theta_T
            data_sim_res$n_max <- n_max
            data_sim_res$n_sim <- i
            data_sim_res$safety_threshold <- safety_threshold
  
            if (data_sim_res$pp<=safety_threshold) {
              n <- n+n_increase
              data_core <- c(data_core, rbinom(n_increase, size=1, prob=true_p0))
              
            }
            else {
              n <- n_max
              n <- n+n_increase
            }
            res_sim[[i]] <- data_sim_res
          }
        }
        res_overall1 <- bind_rows(res_overall1, res_sim)
      }
    }
  }
}

res_sim <- c()
res_overall2 <- c()

for (true_p0 in true_p_0_values) {
  for (theta_T in theta_T_values) {
    for (n_max in n_max_values) {
      for (i in 1:n_sim) {
  
        data_core <- rbinom(initial_n, size=1, prob=true_p0)
        
        n <- initial_n
        
        while (n<(n_max+1)) {
          
          data_sim <- data.frame(n=length(data_core), r=sum(data_core))
          data_sim$postdist <- 1-pbeta(p_0_max, a_0+data_sim$r, b_0+data_sim$n-data_sim$r)
          data_sim$stop <- ifelse(data_sim$postdist>theta_T, 1, 0)
          data_sim$true_p0 <- true_p0
          data_sim$theta_T <- theta_T
          data_sim$n_max <- n_max
          data_sim$n_sim <- i
          

          if (data_sim$stop==0) {
            n <- n+n_increase
            data_core <- c(data_core, rbinom(n_increase, size=1, prob=true_p0))
            
          }
          else {
            n <- n_max
            n <- n+n_increase
          }
          res_sim[[i]] <- data_sim
        }
      }

      res_overall2 <- bind_rows(res_overall2, res_sim)
    }
  }
}

res_overall2$pp <- res_overall2$postdist
res_overall2$safety_threshold <- res_overall2$theta_T
res_overall2$ind <- 1

res_overall <- bind_rows(res_overall1, res_overall2)
```

```{r}
#| echo: false
#| warning: false
#| cache: false
#| fig-height: 10
#| fig-width: 8

res_overall <- res_overall %>% mutate(stop_safety=ifelse(pp>safety_threshold, 1, 0), stop_tox=ifelse(pp>safety_threshold, 1, 0))
res_overall$safety_threshold[res_overall$ind%in%1] <- 999
# Trials who reach N_max cannot stop early
res_overall$stop_safety[res_overall$n==res_overall$n_max] <- 0

res_overall$obs_tox <- round(res_overall$r/res_overall$n, 3)

res_agg <- res_overall %>%  group_by(n_max, theta_T, true_p0, safety_threshold) %>% summarise(stop_safety=mean(stop_safety), stop_tox=mean(stop_tox), obs_tox=mean(obs_tox), exp_n=mean(n), exp_r=mean(r))
res_agg$theta_T <- factor(res_agg$theta_T, levels=c(0.75,0.8,0.9), labels=paste0("Theta_T: ", c(0.75,0.8,0.9)))
res_agg$true_p0 <- factor(res_agg$true_p0, levels=c(0.2,0.25,0.3), labels=c(paste0("True p=", c(0.2), " (H0)"), paste0("True p=", c(0.25,0.3), " (H1)")))
res_agg$safety_threshold <- factor(res_agg$safety_threshold, levels=c(0.8,0.9,1, 999), labels=c(paste0("Safety threshold: ", c(0.8,0.9,1)), "Posterior distribution"))

res_agg$n_max <- as.factor(res_agg$n_max)

fig <- ggplot(res_agg, aes(x=n_max, y=stop_safety, shape=safety_threshold)) + geom_point(position=position_dodge2(width = 0.2)) + facet_wrap(~true_p0+theta_T, ncol=2) + theme_bw() + theme(panel.grid.minor = element_blank())+geom_hline(yintercept=0.2, linetype="dashed") + geom_hline(yintercept=0.8, linetype="dashed") + scale_y_continuous(breaks=c(0,0.2,0.5,0.8, 1), limits=c(0,1))+xlab("N_max")+ylab("Probability to reject") + scale_shape("Stopping decision rule")+ggtitle("Probability to reject")
fig

fig <- ggplot(res_agg, aes(x=n_max, y=stop_tox, shape=safety_threshold)) + geom_point(position=position_dodge2(width = 0.2)) + facet_wrap(~true_p0+theta_T, ncol=2) + theme_bw() + theme(panel.grid.minor = element_blank())+geom_hline(yintercept=0.2, linetype="dashed") + geom_hline(yintercept=0.8, linetype="dashed") + scale_y_continuous(breaks=c(0,0.2,0.5,0.8, 1), limits=c(0,1))+xlab("N_max")+ylab("Probability") + scale_shape("Stopping decision rule")+ggtitle("Probability for declaring toxicity")
fig
```

```{r}
#| echo: false
#| warning: false
#| cache: false
#| fig-height: 5
#| fig-width: 8

fig <- ggplot(res_agg %>% filter(true_p0=="True p=0.2 (H0)"), aes(x=n_max, y=exp_n, shape=safety_threshold)) + geom_point(position=position_dodge2(width = 0.2)) + facet_wrap(~theta_T) + theme_bw() + theme(panel.grid.minor = element_blank())+xlab("N_max")+ylab("Expected N under H0") + scale_shape("Stopping decision rule")
fig+ggtitle("Expected N under H0")

# library(DT)
# datatable(res_agg, rownames = F, filter = 'top')
```

## Randomised non-comparitive two arm design with adaptive excessive toxicity treshold

Suppose we randomise patients with a 1:1 allocation ratio to an experimental and a control arm. The $j$th interim analysis will be performed after the recruitment of $N_j=n_{j,0}+n_{j,1}$ patients ($n_{j,0}$ denotes the number of patients in the control arm and $n_{j,1}$ denotes the number of patients in the experimental arm). For notational simplicity we skip the index $j$ in the following.

Suppose that at an interim analysis **in the experimental arm** $r_{1}$ toxicities are observed. If we assume that $N_{max}=n_{max, 0}+n_{max, 1}$ is the planned sample size of the trial, then a remaining future $m_{1}=n_{max, 1}-n_{1}$ will be recruited in the experimental arm with possible future events $S_1\in\{0,\cdots, m_1\}$.

**The trial is stopped early** if $PP_{1}>\theta_S$, where
$$
PP_{1}=\sum_{0\leq s_1\leq m_{1}} P(S_{1}=s_1|r_{1})\cdot I(T_{i,1}>\theta_T),
$$
and $T_{i,1}:=P(p_1 > p_{max}|r_{1},n_1,S_{1}=i)$. If $PP_{1}\leq \theta_S$ then $n^*$ new patients are randomised.

At each interim step, **we allow the threshold toxicity rate to be adapted**. Let

$$
\begin{aligned}
PP_{diff}&=E\left[I\left\{ P\left(p_{0}-p_{1}> \delta|r_{0}, r_{1}, n_{0}, n_{1}, S_{0}, S_{1}\right)>\omega\right]|r_{0}, r_{1}, n_{0}, n_{1}\right\} \\
&=\sum_{0\leq s_0 \leq m_0}\sum_{0\leq s_1 \leq m_1}I\left\{ P\left(p_{0}-p_{1}> \delta|r_{0}, r_{1}, n_{0}, n_{1}, S_{0}, S_{1}\right)>\omega\right\} \\
&\quad \quad \cdot P(S_0=s_0|r_0, n_0)P(S_1=s_1|r_1, n_1),
\end{aligned}
$$
be the predictive distribution of a relevant toxicity difference at the end of the trial @Dmitrienko2006. $S_i$ are betabinomial distributed $P(S_i=s_i|r_i,n_i) \sim betabinomial(m_i, a_i+r_i, b_i-n_{max,i}-r_i)$. Note that

$$
\begin{aligned}
P&\left(p_{0}-p_{1}> \delta|r_{0}, r_{1}, n_{0}, n_{1}, s_{0}, s_{1}\right)=\int_{\delta}^1\int_0^{p-\delta} f(p_0, p_1|r_0, n_0, r_1, n_1, s_0, s_1)dp_0dp_1 \\
&=\int_{\delta}^1\frac{p^{a_0+r_0+s_0-1}(1-p)^{b_0+n_{0,max}-r_0-s_0-1}}{B(a_0+r_0+s_0, b_0+n_{0,max}-r_0-s_0)}\int_0^{p-\delta}\frac{q^{a_1+r_1+s_1-1}(1-q)^{b_1+n_{1,max}-r_1-s_1-1}}{B(a_1+r_1+s_1, b_1+n_{1,max}-r_1-s_1)}dp \\
&=\int_{\delta}^1\frac{p^{a_0+r_0+s_0-1}(1-p)^{b_0+n_{0,max}-r_0-s_0-1}}{B(a_0+r_0+s_0, b_0+n_{0,max}-r_0-s_0)}F_{B(a_1+r_1+s_1,b_1+n_1-r_1-s_1)}(p-\delta^*)dp,
\end{aligned}
$$
where $F_{B(a_1+r_1+s_1,b_1+n_1-r_1-s_1)}$ is the cumulative distribution function of a beta distribution.

If $PP_{diff}>\theta_D$ then $p_{max}$ is reduced by $d_{max}$ until a lower bound $p_{max}^l$ is reached. If $PP_{diff}\leq \theta_D$, then $p_{max}$ remains unchanged.

```{r fig1}
#| echo: false
#| warning: false
#| results: hide
#| fig-height: 8
#| fig-width: 8

library(Gmisc)
library(grid)

grid.newpage()

x_coord1 <- 0.5

text_box1 <- "1:1 randomisation"
box1 <- boxGrob(text_box1, x=0.5, y=0.9, bjust = "center")
box1

text_box2 <- "Experimental group\n(with assumed true p1)"
box2 <- boxGrob(text_box2, x=0.2, y=0.6, bjust = "center")
box2

text_box3 <- "Control group\n(with assumed true p0)"
box3 <- boxGrob(text_box3, x=0.8, y=0.6, bjust = "center")
box3

text_box4 <- "Interim analysis 1,...,u\n\nStop early if experimental group has excessive toxicity rate\notherwise randomise new patients until N_max is reached\n\nDiscontinuation threshold adaption:\nIf predictive probability that the experimental arm has lower toxicity rate\nthan control arm exceeds prespecified value,\nthen reduce discontinuation threshold by d_max"
box4 <- boxGrob(text_box4, x=0.5, y=0.2, bjust = "center")
box4

connectGrob(box1, box2, "vertical")
connectGrob(box1, box3, "vertical")
```

### Simulation study

We set up a simulation study ("Threshold adaption") to quantify operations characteristics

- Probability of early stopping (stopping trial before $N_{max}$ is reached),
- Probability of toxicity (declaring toxicity including $N_{max}$),
- Probability of efficacy (high probability that experimental arm has a lower toxcity by at least $\delta$ compared to control arm)
- Average (adapted) maximal toxicity threshold
- Expected number of patients per arm

with the following parameters

- Beta prior parameters: Control arm: a=2, b=8; Experimental arm, a=1, b=4
- True control $p0$: 0.1, 0.2, 0.3
- True experimental $p1$: 0.1, 0.2, 0.3
- Maximal toxicity threshold $p_{max}$: 0.2
- Reduction of maximal toxicity threshold $d_{max}$: 0.02
- Lower bound of maximal toxicity threshold $p_{max}^l$: 0.1
- Sample size at first interim analysis $N_1$: 8
- Increase of sample size $n_{increase}$: 4 (2 patients per arm)
- Planned sample size $N_{max}$: 40, 48, 56 (20, 24, 28 patients per arm)
- Excessive toxicity probability for posterior distribution $\theta_T$: 0.8
- Safety threshold for predictive distribution $\theta_S$: 0.8
- Toxicity difference margin $\delta$: 0.025
- Number of simulation runs: 1000

A second simulation studies ("No threshold adaption") does not adapt the discontinuation threshold at interim analyses.

```{r}
#| echo: true
#| warning: false

# Prior information control
a_0 <- 2
b_0 <- 8

x <- seq(0.001,0.999,0.001)

data_prior <- data.frame(x, y=dbeta(x, a_0, b_0), type=1)

# Prior information experimental
a_0 <- 1
b_0 <- 4

data_prior <- bind_rows(data_prior, data.frame(x, y=dbeta(x, a_0, b_0), type=2))

data_prior$type <- factor(data_prior$type, levels=1:2, 
                          labels=c("a=2, b=8 (Control arm)", "a=1, b=4 (Experimental arm)"))
ggplot(data_prior %>% filter(y<5), aes(x=x, y=y, linetype=type, group=type))+
  geom_line()+theme_bw()+scale_x_continuous(breaks=seq(0,1,0.2))+
  theme(panel.grid = element_blank(), legend.position = "bottom")+
  scale_linetype("Prior parameters")+geom_vline(xintercept=0.2, colour="red", linetype=1)+
  geom_vline(xintercept=0.2, colour="red", linetype=2)+
  ylab("Density")+xlab("")+ggtitle("Beta prior parameter choices for simulation study")+
  labs(caption=c("Red vertical lines indicate mean values of priors."))
```

```{r}
#| echo: false
#| warning: false
#| cache: true

library(blockrand)

set.seed(52359)

n_sim <- 1000

a_0 <- 2
b_0 <- 8

a_1 <- 1
b_1 <- 4

p_max <- 0.2
p_max_reduction <- 0.02
p_max_min <- 0.12

true_pmax_0_values <- c(0.1, 0.2, 0.3)
true_pmax_1_values <- c(0.1, 0.2, 0.3)

n_max_values <- c(22, 24, 26)
theta_T_values <- c(0.8)
safety_threshold_values <- c(0.8)
initial_n <- 8
n_increase <- 4

delta_star_values <- c(0.025)
post_delta_threshold <- 0.8
pp_diff_threshold <- 0.8

res_sim <- c()
res_overall1 <- c()

for (delta_star in delta_star_values) {
  for (safety_threshold in safety_threshold_values) {
    for (true_p0_control in true_pmax_0_values) {
      for (true_p0_experimental in true_pmax_1_values) {
        for (theta_T in theta_T_values) {
          for (n_max in n_max_values) {
            for (i in 1:n_sim) {
              # Block randomise patients to arms
              alloc <- as.numeric(blockrand(n=initial_n, block.sizes = 1, levels=0:1)$treatment)-1
              data_control <- rbinom(sum(alloc==0), size=1, prob=true_p0_control)
              data_experimental <- rbinom(sum(alloc==1), size=1, prob=true_p0_experimental)
              n <- length(data_control)+length(data_experimental)

              p_max0 <- p_max
              while (n<(n_max*2+1)) {
  
                data_sim_res <- data.frame(n_0=length(data_control), r_0=sum(data_control), n_1=length(data_experimental), r_1=sum(data_experimental), p_max0=p_max0)

                # Block for PP_diff
                data_sim1 <- expand.grid(i_1=0:(n_max-data_sim_res$n_1), i_0=0:(n_max-data_sim_res$n_0))
                data_sim1$r_1 <- data_sim_res$r_1
                data_sim1$r_0 <- data_sim_res$r_0

                data_sim1$n_1 <- data_sim_res$n_1
                data_sim1$n_0 <- data_sim_res$n_0
                data_sim1$m_1 <- n_max-data_sim_res$n_1
                data_sim1$m_0 <- n_max-data_sim_res$n_0

                data_sim1$joint_prob <- dbbinom(data_sim1$i_1, size=data_sim1$m_1, alpha=a_1+data_sim1$r_1, beta=b_1+data_sim1$n_1-data_sim1$r_1)*dbbinom(data_sim1$i_0, size=data_sim1$m_0, alpha=a_0+data_sim1$r_0, beta=b_0+data_sim1$n_0-data_sim1$r_0)
                
                 for (j in 1:nrow(data_sim1)) { 
                data_sim1$post_delta[j] <- integrate(function(x) (x)^(a_0+data_sim1$r_0[j]+data_sim1$i_0[j]-1)*(1-x)^(b_0+n_max-data_sim1$r_0[j]-data_sim1$i_0[j]-1)*pbeta(x-delta_star, a_1+data_sim1$r_1[j]+data_sim1$i_1[j], b_1+n_max-data_sim1$r_1[j]-data_sim1$i_1[j]), lower = delta_star, upper = 1)[[1]]*(1/beta(a_0+data_sim1$r_0[j]+data_sim1$i_0[j], b_0+n_max-data_sim1$r_0[j]-data_sim1$i_0[j]))
              }
             
           data_sim1$ind <- ifelse(data_sim1$post_delta>post_delta_threshold, 1, 0)
                data_sim1 <- as.data.table(data_sim1)
                data_sim_res$pp_diff <- as.numeric(data_sim1[, .(sum(joint_prob*ind))])
                data_sim_res$delta_star <- delta_star
                
                data_sim <- data.frame(n=rep(length(data_experimental), n_max-length(data_experimental)+1), r=rep(sum(data_experimental), n_max-length(data_experimental)+1))
                data_sim$i <- (1:nrow(data_sim))-1
                data_sim$m <- n_max-data_sim$n
                data_sim$cond_prob <- dbbinom(data_sim$i, size=data_sim$m, alpha=a_1+data_sim$r, beta=b_1+data_sim$n-data_sim$r)
                data_sim$ind <- ifelse(1-pbeta(p_max0, a_1+data_sim$r+data_sim$i, b_1+n_max-data_sim$r-data_sim$i)>theta_T, 1, 0)
                data_sim <- as.data.table(data_sim)
                data_sim_res$pp_experimental <- as.numeric(data_sim[, .(sum(cond_prob*ind))])
                data_sim_res$true_p0_experimental <- true_p0_experimental
                data_sim_res$n_max <- n_max
                data_sim_res$theta_T <- theta_T
                data_sim_res$safety_threshold <- safety_threshold
                data_sim_res$true_p0_control <- true_p0_control
                data_sim_res$n_sim <- i

                 if (data_sim_res$pp_experimental<=safety_threshold) {

                                    if (data_sim_res$pp_diff>pp_diff_threshold & p_max0>p_max_min & data_sim_res$n_0<data_sim_res$n_max) p_max0 <- p_max0-p_max_reduction
                             
                  
                            alloc <- as.numeric(blockrand(n=n_increase, block.sizes = 1, levels=0:1)$treatment)-1
              
                  data_control <- c(data_control, rbinom(sum(alloc==0), size=1, prob=true_p0_control))
                  data_experimental <- c(data_experimental, rbinom(sum(alloc==1), size=1, prob=true_p0_experimental))
                  n <- length(data_control)+length(data_experimental)
                                    
    
                  #  
                }
                else {
                  n <- n_max*2
                  n <- n+1
                  # data_sim_res$p_max <- p_max
                }      

            
              }
              
              
                if (data_sim_res$n_0==data_sim_res$n_max) {
                  data_sim_res$pp_diff <- integrate(function(x) (x)^(a_0+data_sim_res$r_0-1)*(1-x)^(b_0+n_max-data_sim_res$r_0-1)*pbeta(x-delta_star, a_1+data_sim_res$r_1, b_1+n_max-data_sim_res$r_1), lower = delta_star, upper = 1)[[1]]*(1/beta(a_0+data_sim_res$r_0, b_0+n_max-data_sim_res$r_0))
                 
                }
            
  res_sim[[i]] <- data_sim_res
            }

        res_overall1 <- bind_rows(res_overall1, res_sim)
    
    }
  }
}
}
  }
}

res_sim <- c()
res_overall2 <- c()

for (delta_star in delta_star_values) {
  for (safety_threshold in safety_threshold_values) {
    for (true_p0_control in true_pmax_0_values) {
      for (true_p0_experimental in true_pmax_1_values) {
        for (theta_T in theta_T_values) {
          for (n_max in n_max_values) {
            for (i in 1:n_sim) {
              alloc <- as.numeric(blockrand(n=initial_n, block.sizes = 1, levels=0:1)$treatment)-1
              data_control <- rbinom(sum(alloc==0), size=1, prob=true_p0_control)
              data_experimental <- rbinom(sum(alloc==1), size=1, prob=true_p0_experimental)
              n <- length(data_control)+length(data_experimental)

              p_max0 <- p_max
              while (n<(n_max*2+1)) {
  
                data_sim_res <- data.frame(n_0=length(data_control), r_0=sum(data_control), n_1=length(data_experimental), r_1=sum(data_experimental), p_max0=p_max0)

                # Block for predictive P(p1-p0>delta_star)
                data_sim1 <- expand.grid(i_1=0:(n_max-data_sim_res$n_1), i_0=0:(n_max-data_sim_res$n_0))
                data_sim1$r_1 <- data_sim_res$r_1
                data_sim1$r_0 <- data_sim_res$r_0

                data_sim1$n_1 <- data_sim_res$n_1
                data_sim1$n_0 <- data_sim_res$n_0
                data_sim1$m_1 <- n_max-data_sim_res$n_1
                data_sim1$m_0 <- n_max-data_sim_res$n_0

                data_sim1$joint_prob <- dbbinom(data_sim1$i_1, size=data_sim1$m_1, alpha=a_1+data_sim1$r_1, beta=b_1+data_sim1$n_1-data_sim1$r_1)*dbbinom(data_sim1$i_0, size=data_sim1$m_0, alpha=a_0+data_sim1$r_0, beta=b_0+data_sim1$n_0-data_sim1$r_0)
                
                 for (j in 1:nrow(data_sim1)) { 
                data_sim1$post_delta[j] <- integrate(function(x) (x)^(a_0+data_sim1$r_0[j]+data_sim1$i_0[j]-1)*(1-x)^(b_0+n_max-data_sim1$r_0[j]-data_sim1$i_0[j]-1)*pbeta(x-delta_star, a_1+data_sim1$r_1[j]+data_sim1$i_1[j], b_1+n_max-data_sim1$r_1[j]-data_sim1$i_1[j]), lower = delta_star, upper = 1)[[1]]*(1/beta(a_0+data_sim1$r_0[j]+data_sim1$i_0[j], b_0+n_max-data_sim1$r_0[j]-data_sim1$i_0[j]))
              }
             
           data_sim1$ind <- ifelse(data_sim1$post_delta>post_delta_threshold, 1, 0)
                data_sim1 <- as.data.table(data_sim1)
                data_sim_res$pp_diff <- as.numeric(data_sim1[, .(sum(joint_prob*ind))])
                data_sim_res$delta_star <- delta_star
                
                data_sim <- data.frame(n=rep(length(data_experimental), n_max-length(data_experimental)+1), r=rep(sum(data_experimental), n_max-length(data_experimental)+1))
                data_sim$i <- (1:nrow(data_sim))-1
                data_sim$m <- n_max-data_sim$n
                data_sim$cond_prob <- dbbinom(data_sim$i, size=data_sim$m, alpha=a_1+data_sim$r, beta=b_1+data_sim$n-data_sim$r)
                data_sim$ind <- ifelse(1-pbeta(p_max0, a_1+data_sim$r+data_sim$i, b_1+n_max-data_sim$r-data_sim$i)>theta_T, 1, 0)
                data_sim <- as.data.table(data_sim)
                data_sim_res$pp_experimental <- as.numeric(data_sim[, .(sum(cond_prob*ind))])
                data_sim_res$true_p0_experimental <- true_p0_experimental
                data_sim_res$n_max <- n_max
                data_sim_res$theta_T <- theta_T
                data_sim_res$safety_threshold <- safety_threshold
                data_sim_res$true_p0_control <- true_p0_control
                data_sim_res$n_sim <- i

                 if (data_sim_res$pp_experimental<=safety_threshold) {

                  
                            alloc <- as.numeric(blockrand(n=n_increase, block.sizes = 1, levels=0:1)$treatment)-1
              
                  data_control <- c(data_control, rbinom(sum(alloc==0), size=1, prob=true_p0_control))
                  data_experimental <- c(data_experimental, rbinom(sum(alloc==1), size=1, prob=true_p0_experimental))
                  n <- length(data_control)+length(data_experimental)
                                    
    
                  #  
                }
                else {
                  n <- n_max*2
                  n <- n+1
                  # data_sim_res$p_max <- p_max
                }      

            
              }
              
              
                if (data_sim_res$n_0==data_sim_res$n_max) {
                  data_sim_res$pp_diff <- integrate(function(x) (x)^(a_0+data_sim_res$r_0-1)*(1-x)^(b_0+n_max-data_sim_res$r_0-1)*pbeta(x-delta_star, a_1+data_sim_res$r_1, b_1+n_max-data_sim_res$r_1), lower = delta_star, upper = 1)[[1]]*(1/beta(a_0+data_sim_res$r_0, b_0+n_max-data_sim_res$r_0))
                 
                }
            
  res_sim[[i]] <- data_sim_res
            }

        res_overall2 <- bind_rows(res_overall2, res_sim)
    
    }
  }
}
}
  }
}
```


```{r}
#| echo: false
#| warning: false
#| cache: false
#| fig-height: 10
#| fig-width: 8

res_overall1$type <- 0
res_overall2$type <- 1
res_overall <- bind_rows(res_overall1, res_overall2)


# res_overall$safety_threshold[res_overall$ind%in%1] <- 999

res_overall$stop_early <- ifelse(res_overall$n_1<res_overall$n_max, 1, 0)
res_overall$toxicity <- ifelse(res_overall$pp_experimental>safety_threshold, 1, 0)

res_overall$efficacy <- ifelse(res_overall$pp_diff>pp_diff_threshold, 1, 0)

res_agg <- res_overall %>% group_by(true_p0_control, true_p0_experimental, n_max, theta_T, safety_threshold, delta_star, type) %>% summarise(n=n(),n_control=mean(n_0), n_experimental=mean(n_1), stop_early=mean(stop_early), toxicity=mean(toxicity), efficacy=mean(efficacy), p_max=mean(p_max0))
res_agg$toxicity_uci <- res_agg$toxicity+qnorm(0.975)*sqrt(res_agg$toxicity*(1-res_agg$toxicity)/res_agg$n)
res_agg$toxicity_lci <- res_agg$toxicity-qnorm(0.975)*sqrt(res_agg$toxicity*(1-res_agg$toxicity)/res_agg$n)

res_agg$stop_early_uci <- res_agg$stop_early+qnorm(0.975)*sqrt(res_agg$stop_early*(1-res_agg$stop_early)/res_agg$n)
res_agg$stop_early_lci <- res_agg$stop_early-qnorm(0.975)*sqrt(res_agg$stop_early*(1-res_agg$stop_early)/res_agg$n)


res_agg$delta_star <- as.factor(res_agg$delta_star)
res_agg$theta_T <- factor(res_agg$theta_T, levels=c(0.8), labels=paste0("Theta_T: ", c(0.8)))
res_agg$true_p0_experimental <- factor(res_agg$true_p0_experimental, levels=c(0.1, 0.2,0.3), labels=c(paste0("True experimental p=", c(0.1)), paste0("True experimental p=", c(0.2)), paste0("True experimental p=", c(0.3))))
res_agg$true_p0_control <- factor(res_agg$true_p0_control, levels=c(0.1,0.2,0.3), labels=c(paste0("True control p=", c(0.1)), paste0("True control p=", c(0.2)), paste0("True control p=", c(0.3))))
res_agg$safety_threshold <- factor(res_agg$safety_threshold, levels=c(0.8,0.9,1,999), labels=c(paste0("Safety threshold: ", c(0.8,0.9,1)), "Posterior distribution"))
res_agg$n_max <- as.factor(res_agg$n_max)
res_agg$type <- factor(res_agg$type, levels=0:1, labels=c("Threshold adaption", "No threshold adaption"))

fig <- ggplot(res_agg, aes(x=n_max, y=stop_early, colour=type, label=stop_early)) + geom_point(position=position_dodge2(width = 0.2)) + facet_wrap(~true_p0_experimental+true_p0_control) + theme_bw() + theme(panel.grid.minor = element_blank(), legend.position = "bottom")+xlab("Number of patients per arm")+ylab("Probability of early stopping")+scale_shape("Stopping decision rule")+geom_hline(yintercept=0.2, linetype="dashed") + geom_hline(yintercept=0.8, linetype="dashed")+scale_color_manual("Simulation type", values=c("orange", "purple"))+scale_y_continuous(limits=c(0,1))
fig+ggtitle("Probability of early stopping")

fig <- ggplot(res_agg, aes(x=n_max, y=toxicity, colour=type)) + geom_point(position=position_dodge2(width = 0.2)) + facet_wrap(~true_p0_experimental+true_p0_control) + theme_bw() + theme(panel.grid.minor = element_blank(), legend.position = "bottom")+xlab("Number of patients per arm")+ylab("Probability of toxicity")+scale_shape("Stopping decision rule")+geom_hline(yintercept=0.2, linetype="dashed") + geom_hline(yintercept=0.8, linetype="dashed")+scale_color_manual("Simulation type", values=c("orange", "purple"))+scale_y_continuous(limits=c(0,1))
fig+ggtitle("Probability of toxicity")

fig <- ggplot(res_agg, aes(x=n_max, y=efficacy, colour=type)) + geom_point(position=position_dodge2(width = 0.2)) + facet_wrap(~true_p0_experimental+true_p0_control) + theme_bw() + theme(panel.grid.minor = element_blank(), legend.position = "bottom")+xlab("Number of patients per arm")+ylab("Probability of efficacy")+scale_shape("Stopping decision rule")+geom_hline(yintercept=0.2, linetype="dashed") + geom_hline(yintercept=0.8, linetype="dashed")+scale_color_manual("Simulation type", values=c("orange", "purple"))+scale_y_continuous(limits=c(0,1))
fig+ggtitle("Probability of efficacy")

fig <- ggplot(res_agg, aes(x=n_max, y=p_max, colour=type)) + geom_point(position=position_dodge2(width = 0.2)) + facet_wrap(~true_p0_experimental+true_p0_control) + theme_bw() + theme(panel.grid.minor = element_blank(), legend.position = "bottom")+xlab("Number of patients per arm")+ylab("Toxicity threshold")+scale_shape("Stopping decision rule")+geom_hline(yintercept=0.2, linetype="dashed") + geom_hline(yintercept=0.8, linetype="dashed")+scale_color_manual("Simulation type", values=c("orange", "purple"))+scale_y_continuous(limits=c(0,0.2))
fig+ggtitle("Average toxicity threshold")
```

```{r}
#| echo: false
#| warning: false
#| cache: false
#| fig-height: 5
#| fig-width: 8

fig <- ggplot(res_agg %>% filter(true_p0_control=="True control p=0.2", true_p0_experimental=="True experimental p=0.2"), aes(x=n_max, y=n_control, colour=type)) + geom_point(position=position_dodge2(width = 0.2)) + theme_bw() + theme(panel.grid.minor = element_blank())+xlab("N_max")+ylab("Expected N under H0 per arm")+scale_color_manual("Simulation type", values=c("orange", "purple"))
fig+ggtitle("Expected N under H0")

# library(DT)
# datatable(res_agg, rownames = F, filter = 'top')
```